---
title: "Multilayer perceptron in Keras"
date: "28 March 2022"
output: html_document
---

--------------------------------------------------------------------------------

#### Preliminary remarks

To install `keras` and run the notebooks from this week:

1. Run `install.packages("keras")`
2. Load the package with `library("keras")`
3. Then run the function `install_keras()` in the R-console

Afterwards the code in this notebook should run properly.

For more information on `tensorflow` (the basis of `keras`) see: https://tensorflow.rstudio.com/installation/

--------------------------------------------------------------------------------

Code from: https://github.com/rstudio/keras/blob/master/vignettes/examples/mnist_mlp.R

In this notebook, we will discuss the following example of an MLP classifying the MNIST digits.

Loading packages:

```{r}
library("keras")
```

Data preparation

```{r}
batch_size <- 128
num_classes <- 10
epochs <- 8

# Using the MNIST data which comes with the Keras/Tensorflow packages

# The data, shuffled and split between train and test sets
c(c(x_train, y_train), c(x_test, y_test)) %<-% dataset_mnist()

x_train <- array_reshape(x_train, c(nrow(x_train), 784))
x_test <- array_reshape(x_test, c(nrow(x_test), 784))

# Divide by max value to transform into [0,1] range
x_train <- x_train / 255
x_test <- x_test / 255

cat(nrow(x_train), 'train samples\n')
cat(nrow(x_test), 'test samples\n')

# Convert class vectors to binary class matrices
y_train <- to_categorical(y_train, num_classes)
y_test <- to_categorical(y_test, num_classes)

```

Each observation is a row vector of 784 observations (originally 28x28 pixels, now flattened):

```{r}
dim(x_train)
```

Define the model:

```{r}
model <- keras_model_sequential()
model %>% 
  layer_dense(units = 256, activation = 'relu', input_shape = c(784)) %>% 
  layer_dropout(rate = 0.4) %>% 
  layer_dense(units = 128, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 10, activation = 'softmax')

summary(model)

model %>% compile(
  loss = 'categorical_crossentropy',
  optimizer = optimizer_rmsprop(),
  metrics = c('accuracy')
)
```

Training:

```{r}
history <- model %>% fit(
  x_train, y_train,
  batch_size = batch_size,
  epochs = epochs,
  verbose = 1,
  validation_split = 0.2
)
```

Evaluation:

```{r}
score <- model %>% evaluate(
  x_test, y_test,
  verbose = 0
)
```

Plot:

```{r}
plot(history)
```

Output metrics:

```{r}
cat('Test loss:', score[[1]], '\n')
cat('Test accuracy:', score[[2]], '\n')
```

Training for some more epochs will likely improve the model further. You can also try out some different hyper-parameters here, e.g. dropout or different amounts of hidden layers, etc.


References:

  - https://github.com/rstudio/keras/blob/master/vignettes/examples/mnist_mlp.R